Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-07 MS/ms.Rmd
Writing heterogeneity section. Trying to link back to initial point of paper a bit better

## **Introduction**
Meta-analysis provides quantitatively informed answers to research questions by using a transparent and reproducible search strategy to synthesise research results across studies and identify sources of variation [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide on a given question - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) a meta-analyst can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Quntitative syntheses, including meta-analyses, have provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer orgamisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bisphenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominant the field of comparative physiology. Formal meta-analyses involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMAEcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error [@Koricheva2013]. The inverse sampling error is used to weight higher powered studies in the analysis and make overall esatimtes more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased.

Despite the widespread adoption and well established methodological procedures of meta-analysis it is often criticised for mixing 'apples and oranges' - in otherwords, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widspread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. In addition, they often include a diversity of study species which can vary in responses.

@@

#### *Multi-level Meta-Analysis when the Overall Effect is of Interest*
One of the main goals of meta-analysis is to obtain an overall meta-analytic mean estimate (and some form of uncertainty around that mean, like a 95% confidence interval, may be even with a prediction interval - @Nakagawa2021c) which describes the magnitude of effect across all the studies / effects included in the meta-analysis. This is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierachical structure. For example, multiple effect sizes may be taken from a single study on different traits measured on the same individuals, or come from many different species that share an evolutionary history [@Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multi-level meta-analytic model to control for the various sources of non-independence [@NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance. For the purpose of our discussion, we will assume the meta-analyst plans to use more advanced multilevel models. Having said that, it maybe sufficient for many meta-analyses to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

#### *Heterogenity Analysis: How Much Effect Size Variation Actually Exists?*
Quantifying effect size heterogenity (i.e., effect size variance) provides a pathway to explore hypotheses thought to explain why effects might vary within and across studies. Effect size heterogeneity can be thought of as the percentage of variance between effect sizes that cannot be attributable to sampling error (often as called 'total' heterogenity) @Higgins2002. It provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance). We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure becasue only then can we quantify how much variation is the result of sampling variability. @Senior2016 has shown that total heterogenity is often extremely high in ecological and evolutionary meta-analysis (~91%), suggesting that effect measures vary widely across within and studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies

>>>>

## **Introduction**
Meta-analysis provides quantitatively informed answers to research questions by using a transparent and reproducible search strategy to synthesise research results across studies and identify sources of variation [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide on a question - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) a meta-analyst can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Quntitative syntheses, including meta-analyses, have provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer orgamisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bisphenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominant the field of comparative physiology. Formal meta-analyses involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. The entire process should follow and report upon this entire process carefully to ensure the meta-analysis is as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMAEcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of the research question, along with identifying any gaps and biases, there are two major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship and 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013]. A critically important feature of meta-analaysis is its ability to account for sampling error [@Koricheva2013]. The inverse sampling error is used to weight higher powered studies in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased.

Despite the widespread adoption and well established methodological procedures of meta-analysis it is often criticised for mixing 'apples and oranges' - in otherwords, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widspread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. In addition, they often include a diversity of study species which can vary in responses. To the uninitiated the concern might not be unfounded given that meta-analyses in ecology and evolution often mix highly heterogenous effect size estimates often resulting in high variability [@Senior2016].

@@

#### *Multi-level Meta-Analysis when the Overall Effect is of Interest*
One of the main goals of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be quite useful because, in the context of a meta-analysis, it provides the range of effect size values we might expect to observe if a new study were to be conducted [@Nakagawa2021c]. Estimating an overall mean along with 95% CI (or PI) is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierachical structure. For example, multiple effect sizes may be taken from a single study on different traits measured on the same individuals, or come from many different species that share an evolutionary history [@Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multi-level meta-analytic model to control for the various sources of non-independence [@NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance. For the purpose of our discussion, we will assume the meta-analyst plans to use more advanced multilevel models. Having said that, it maybe sufficient for many meta-analyses to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

#### *Heterogenity Analysis: How Much Effect Size Variation Actually Exists?*
Quantifying effect size heterogenity (i.e., effect size variance) provides a pathway to explore hypotheses thought to explain why effects might vary within and across studies. Effect size heterogeneity can be thought of as the percentage of variance between effect sizes that cannot be attributable to sampling error (often as called 'total' heterogenity) [@Higgins2002]. More formally, we can quantify heterogenity using the multilevel meta-analytic model we describe above as:

$$ I^2_{total} = \frac{\sigma^2_{total} - \sigma^2_{m}}{\sigma^2_{total}} $$
where $\sigma^2_{total}$ is the  total effect size variance (using the above multilevel model than $\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}$) and $\sigma^2_{m}$ is the total sampling error variance.  Quantifying heterogenity provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure because only then can we quantify how much variation is the result of sampling variability. @Senior2016 have shown that total heterogenity is often extremely high in ecological and evolutionary meta-analysis (~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogenity estimates from meta-analytic models can provide a way to undersand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012:

$$ I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}} $$
where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects.

** NOTE THAT WE MAY WNAT TO SWITCH POSITION, INTRODUCE METEREGRESSION FIRSTs
